---
title: "Model Formulation and Assumptions in the Linear Model"
subtitle: "Growth data, diagnostic plots, and the LINE assumptions"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, fig.align = "center")
```

## Before you start

This notebook rebuilds every figure from the lecture, using code you can read,
run, and change. Each section ends with a "**Try it yourself**" exercise, make sure to
do those before moving on. They are where the actual learning happens, not
the code above them.

You will need the following four packages. Run this once:

```{r install, eval=FALSE}
install.packages(c("ggplot2", "dplyr", "broom", "lmtest"))
```

```{r libraries}
library(ggplot2)
library(dplyr)
library(broom)
library(lmtest)

set.seed(2026)

NAVY   <- "#1E2761"
ACCENT <- "#E8871E"
GREY   <- "#898781"
ICE    <- "#CADCFC"
GREEN  <- "#1D9E75"

theme_deck <- theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(),
        axis.title = element_text(face = "bold"))
```

---

## 1. The growth data

We anchor everything to **real CDC growth-chart medians** (Source: https://www.cdc.gov/growthcharts/cdc-data-files.htm). 
The curve in all figures are pulled from the real CDC data and the individual data points alongside those curves 
represent the simulated data points.

Rather than typing in a handful of milestone values by hand, the following chunk of 
code pulls the **exact published medians directly from CDC's own data files**
every time it runs. (Alternatively, you could download those files locally and upload them in your RStudio environment, 
instead of fetching them from the CDC website.)
Note that two files per measurement are needed, because CDC publishes infants
(birth–36 months) and older children (2–20 years) as separate tables.

```{r cdc-anchors}
# Official CDC/NCHS data files (2000 growth charts), fetched live.
# Source: https://www.cdc.gov/growthcharts/cdc-data-files.htm
statage  <- read.csv("https://www.cdc.gov/growthcharts/data/zscore/statage.csv")   # stature, 2-20 yr
wtage    <- read.csv("https://www.cdc.gov/growthcharts/data/zscore/wtage.csv")     # weight, 2-20 yr
lenageinf<- read.csv("https://www.cdc.gov/growthcharts/data/zscore/lenageinf.csv") # length, birth-36 mo
wtageinf <- read.csv("https://www.cdc.gov/growthcharts/data/zscore/wtageinf.csv")  # weight, birth-36 mo

# CDC's files sometimes read Agemos/M as character (trailing whitespaces
# or a stray footnote character can do this). CDC also concatenates
# the boys' and girls' tables in one CSV, with a second header row at
# the transition -- read.csv() does not catch it, so it becomes a data
# row full of NAs once converted. We drop those rows explicitly below.
fix_numeric <- function(df) {
  df$Agemos <- as.numeric(trimws(df$Agemos))
  df$M      <- as.numeric(trimws(df$M))
  df$Sex    <- as.numeric(trimws(df$Sex))
  df[!is.na(df$Sex) & !is.na(df$Agemos) & !is.na(df$M), ]
}
statage   <- fix_numeric(statage)
wtage     <- fix_numeric(wtage)
lenageinf <- fix_numeric(lenageinf)
wtageinf  <- fix_numeric(wtageinf)

# Confirm nothing unexpected remains -- all four should print 0 now
cat("NAs remaining -- statage:", sum(is.na(statage$Agemos)) + sum(is.na(statage$M)),
    "| wtage:", sum(is.na(wtage$Agemos)) + sum(is.na(wtage$M)),
    "| lenageinf:", sum(is.na(lenageinf$Agemos)) + sum(is.na(lenageinf$M)),
    "| wtageinf:", sum(is.na(wtageinf$Agemos)) + sum(is.na(wtageinf$M)), "\n")

# Sex == 1 is male in all four files. We use the boys' curve throughout
# for consistency. 
# The same code works for Sex == 2 (female), as an exercise you could re-run this code 
# to compare males to females
boys_height_young <- lenageinf[lenageinf$Sex == 1, c("Agemos", "M")]
boys_height_older <- statage[statage$Sex == 1, c("Agemos", "M")]
boys_weight_young <- wtageinf[wtageinf$Sex == 1, c("Agemos", "M")]
boys_weight_older <- wtage[wtage$Sex == 1, c("Agemos", "M")]

# Infant tables cover birth-36mo; switch to the 2-20yr tables from 24mo
# onward, to avoid double-counting the overlapping 24-36mo range.
# NB: Agemos is now guaranteed numeric, so these comparisons are safe.
height_table <- rbind(boys_height_young[boys_height_young$Agemos < 24, ],
                      boys_height_older[boys_height_older$Agemos >= 24, ])
weight_table <- rbind(boys_weight_young[boys_weight_young$Agemos < 24, ],
                      boys_weight_older[boys_weight_older$Agemos >= 24, ])

age_anchor    <- height_table$Agemos / 12   # convert months to years
height_median <- height_table$M
weight_age    <- weight_table$Agemos / 12
weight_median <- weight_table$M

age_grid      <- seq(0, 18, length.out = 200)
height_curve  <- approx(age_anchor, height_median, xout = age_grid)$y
weight_curve  <- approx(weight_age, weight_median, xout = age_grid)$y

cat("Median height at age 5:", round(approx(age_anchor, height_median, xout = 5)$y, 1), "cm\n")
cat("Median height at age 12:", round(approx(age_anchor, height_median, xout = 12)$y, 1), "cm\n")
cat("Median weight at age 10:", round(approx(weight_age, weight_median, xout = 10)$y, 1), "kg\n")
```

> **NB:** if you are offline, or CDC's server is briefly unreachable,
> this chunk will fail — that is expected for a live fetch. The rest of
> the notebook depends on `age_anchor`, `height_median`, `weight_age`,
> and `weight_median` existing, so re-run this chunk once you are back
> online before continuing.

```{r simulate-class}
n_kids     <- 70
ages_sim   <- runif(n_kids, 0, 18)
height_sim <- approx(age_anchor, height_median, xout = ages_sim)$y +
  rnorm(n_kids, 0, 3.5)

growth_data <- data.frame(age = ages_sim, height = height_sim)
head(growth_data)
```

> **Try it yourself:** change `n_kids` to 20, then to 300. Re-run the chunk
> and look at the scatter below. What changes about how clearly you can
> see the underlying trend ? What *does not* change ?

---

## 2. Fitting the model

```{r fit}
fit <- lm(height ~ age, data = growth_data)
summary(fit)
```

`summary()` gives you the estimate, standard error, t-statistic, and
p-value for the slope — this is the output the "inference" half of the
lecture was about. `augment()` (from **broom** R package) adds `.fitted` and
`.resid` columns automatically, which every diagnostic plot below needs:

```{r augment}
growth_aug <- augment(fit)
head(growth_aug)
```

---

## 3. Figure: the linearity reality check

```{r fig1-linearity, fig.width=6, fig.height=4}
ggplot() +
  geom_point(data = growth_data, aes(age, height), color = NAVY, alpha = 0.75, size = 2) +
  geom_line(data = data.frame(age = age_grid, height = height_curve),
            aes(age, height), color = GREEN, linewidth = 1.1) +
  geom_smooth(data = growth_data, aes(age, height), method = "lm",
              se = FALSE, color = ACCENT, linetype = "dashed", linewidth = 1) +
  labs(x = "age (years)", y = "height (cm)",
       title = "Real CDC growth curve vs. a straight-line fit") +
  theme_deck
```

> **Try it yourself:** the straight-line fit (orange dashed) misses both
> ends of the real curve (green). Try fitting `height ~ age + I(age^2)`
> instead — does the fitted curve track the real one more closely? This
> is exactly the "richer X" fix named on the L slide.

---

## 4. Figure: the independence reality check

```{r fig2-independence, fig.width=6, fig.height=4}
ages_track    <- c(2, 4, 6, 8, 10, 12, 14, 16)
child_offsets <- c(-9, -6, -3, 3, 6, 9)   # 3 pairs, above/below median

independence_data <- lapply(seq_along(child_offsets), function(i) {
  base <- approx(age_anchor, height_median, xout = ages_track)$y + child_offsets[i]
  data.frame(
    child  = paste0("child_", i),
    pair   = ceiling(i / 2),
    age    = ages_track,
    height = base + rnorm(length(ages_track), 0, 1.8)
  )
}) %>% bind_rows()

ggplot(independence_data, aes(age, height, group = child, color = factor(pair))) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c(NAVY, ACCENT, GREEN), name = "Child pair") +
  labs(x = "age (years)", y = "height (cm)",
       title = "Repeated measurements on the same children") +
  theme_deck
```

> **Try it yourself:** each "pair" shares a colour. Pick one pair and trace
> just their two lines. Do they track each other more closely than they
> track a different-coloured pair? That visual clustering is the
> independence violation, before any formal test.

---

## 5. Figure: normality / equal-variance reality check (weight)

```{r fig3-variance, fig.width=6, fig.height=4}
n_w        <- 90
ages_w     <- runif(n_w, 0.2, 18)
mean_w     <- approx(weight_age, weight_median, xout = ages_w)$y
weight_sim <- mean_w * exp(rnorm(n_w, 0, 0.18))  # lognormal-ish: skewed, variance grows with mean

weight_data <- data.frame(age = ages_w, weight = weight_sim, mean_w = mean_w)
weight_data$resid <- weight_data$weight - weight_data$mean_w

var_young <- var(weight_data$resid[weight_data$age < 6])
var_old   <- var(weight_data$resid[weight_data$age >= 12])
cat("var(residual), age < 6: ", round(var_young, 1),
    "\nvar(residual), age >= 12:", round(var_old, 1), "\n")
```

```{r fig3-plot, fig.width=6, fig.height=4}
ggplot(weight_data, aes(age, weight)) +
  geom_point(color = NAVY, alpha = 0.75, size = 2) +
  geom_line(data = data.frame(age = age_grid, weight = weight_curve),
            aes(age, weight), color = GREEN, linewidth = 1.1) +
  geom_ribbon(data = data.frame(age = age_grid, weight = weight_curve),
              aes(age, ymin = weight - 4, ymax = weight + 4),
              fill = ICE, alpha = 0.4, inherit.aes = FALSE) +
  labs(x = "age (years)", y = "weight (kg)",
       title = "Real CDC weight curve vs. assumed constant-width band") +
  theme_deck
```

> **Try it yourself:** the printed variances above should show the older
> group's variance is roughly 20 times the younger group's. Try
> `log(weight_sim)` instead of raw weight, recompute the two variances.
> Does log-transforming close that gap ? This is the practical reason
> weight gets log-transformed so often in real analyses.

---

## 6. The eight diagnostic mini-plots: well-behaved vs. violated

Same plot type within each pair. The only thing
that should change between "well-behaved" and "violated" is the pattern,
not the axes.

```{r mini-theme}
theme_mini <- theme_minimal(base_size = 14) +
  theme(axis.text = element_blank(), axis.ticks = element_blank(),
        panel.grid = element_blank(), axis.title = element_text(face = "bold"))
```

### L: Linear (residual vs. fitted)

```{r mini-L, fig.width=8, fig.height=3.2}
n <- 220
fitted_vals <- seq(0, 10, length.out = n)

L_good <- ggplot(data.frame(fitted = fitted_vals, resid = rnorm(n, 0, 1.3)),
                  aes(fitted, resid)) +
  geom_point(color = NAVY, alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE, color = ACCENT, linewidth = 1.3) +
  geom_hline(yintercept = 0, linetype = "dotted", color = GREY) +
  labs(x = "fitted value", y = "residual", title = "Well-behaved") + theme_mini

resid_bad_L <- 0.25 * (fitted_vals - 5)^2 - 3 + rnorm(n, 0, 1.4)
L_bad <- ggplot(data.frame(fitted = fitted_vals, resid = resid_bad_L),
                 aes(fitted, resid)) +
  geom_point(color = NAVY, alpha = 0.55) +
  geom_smooth(method = "loess", se = FALSE, color = ACCENT, linewidth = 1.3) +
  geom_hline(yintercept = 0, linetype = "dotted", color = GREY) +
  labs(x = "fitted value", y = "residual", title = "Violated") + theme_mini

library(patchwork)  # install.packages("patchwork") if needed
L_good | L_bad
```

### I: Independent (residual vs. order of observation)

```{r mini-I, fig.width=8, fig.height=3.2}
order <- 1:60

I_good <- ggplot(data.frame(order = order, resid = rnorm(60, 0, 1)), aes(order, resid)) +
  geom_line(color = NAVY, alpha = 0.6) + geom_point(color = NAVY) +
  geom_hline(yintercept = 0, linetype = "dashed", color = GREY) +
  labs(x = "order of observation", y = "residual", title = "Well-behaved") + theme_mini

ar <- numeric(60); ar[1] <- rnorm(1)
for (t in 2:60) ar[t] <- 0.85 * ar[t - 1] + rnorm(1, 0, 0.6)
I_bad <- ggplot(data.frame(order = order, resid = ar), aes(order, resid)) +
  geom_line(color = NAVY, alpha = 0.6) + geom_point(color = NAVY) +
  geom_hline(yintercept = 0, linetype = "dashed", color = GREY) +
  labs(x = "order of observation", y = "residual", title = "Violated") + theme_mini

I_good | I_bad
```

### N: Normal (QQ-plot)

```{r mini-N, fig.width=8, fig.height=3.2}
N_good <- ggplot(data.frame(resid = rnorm(120)), aes(sample = resid)) +
  stat_qq(color = NAVY, alpha = 0.75) +
  stat_qq_line(color = ACCENT, linewidth = 1.2) +
  labs(x = "theoretical quantiles", y = "sample quantiles", title = "Well-behaved") + theme_mini

N_bad <- ggplot(data.frame(resid = rt(120, df = 3)), aes(sample = resid)) +
  stat_qq(color = NAVY, alpha = 0.75) +
  stat_qq_line(color = ACCENT, linewidth = 1.2) +
  labs(x = "theoretical quantiles", y = "sample quantiles", title = "Violated") + theme_mini

N_good | N_bad
```

### E: Equal variance (residual vs. fitted)

```{r mini-E, fig.width=8, fig.height=3.2}
fitted_e <- runif(160, 1, 9)

E_good <- ggplot(data.frame(fitted = fitted_e, resid = rnorm(160, 0, 1)), aes(fitted, resid)) +
  geom_point(color = NAVY, alpha = 0.7) +
  geom_hline(yintercept = 0, color = ACCENT, linewidth = 1.2) +
  geom_ribbon(data = data.frame(fitted = c(0, 10)), aes(x = fitted, ymin = -2, ymax = 2),
              fill = ICE, alpha = 0.4, inherit.aes = FALSE) +
  labs(x = "fitted value", y = "residual", title = "Well-behaved") + theme_mini

E_bad <- ggplot(data.frame(fitted = fitted_e, resid = rnorm(160, 0, 0.3 + 0.35 * fitted_e)),
                 aes(fitted, resid)) +
  geom_point(color = NAVY, alpha = 0.7) +
  geom_hline(yintercept = 0, color = ACCENT, linewidth = 1.2) +
  labs(x = "fitted value", y = "residual", title = "Violated") + theme_mini

E_good | E_bad
```

> **Try it yourself:** in the `L_bad` chunk, change `0.25 * (fitted_vals - 5)^2`
> to `0.6 * (fitted_vals - 5)^2` (a stronger curvature) and re-run. How much
> more obvious does the U-shape need to get before you would categorise it as
> "violated", rather than borderline ? There is no hard cutoff here.

---

## 7. Formal tests

Every plot above has a matching formal test. Run them on the real fitted
model (`fit`, from Section 2):

```{r formal-tests}
resettest(fit, power = 2:3, type = "fitted")   # linearity
dwtest(fit)                                     # independence
shapiro.test(resid(fit))                        # normality
bptest(fit)                                     # equal variance
```

> **Try it yourself:** for each test, find the p-value in the output.
> A small p-value (conventionally < 0.05) means the test is flagging a
> likely violation. Do the test results agree with what the plots showed
> you visually ? They usually do — but they don't have to, especially
> with a small sample.

---

## Wrap-up

If you have run every chunk of code and every "**Try it yourself**" exercise, you have reproduced
the entire lecture's evidence base yourself using the official CDC/NCHS data !
